Geometric models reveal behavioral and neural signatures of transforming experiences into memories

### 2021-02-19 paper/old.tex

\newcommand{\topicopt}{S1}
\newcommand{\topics}{S2}
\newcommand{\featureimportance}{S3}
\newcommand{\corrmats}{S4}
\newcommand{\matchmats}{S5}
\newcommand{\kopt}{S6}

@@

\begin{abstract}
The mental contexts in which we interpret experiences are often person-specific, even when the experiences themselves are shared. We developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences. We model experiences and memories as \textit{trajectories} through word embedding spaces whose coordinates reflect the universe of thoughts under consideration. Memory encoding can then be modeled as geometrically preserving or distorting the \textit{shape} of the original experience. We applied our approach to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging. Participants’ recountings all preserved coarse spatial properties (essential narrative elements), but not fine spatial scale (low-level) details, of the episode’s trajectory. We also identified networks of brain structures sensitive to these trajectory shapes. Our work provides insights into how we preserve and distort our ongoing experiences when we encode them into episodic memories.
\end{abstract}

\section*{Introduction}
What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete, binary operation: each studied item may be separated from the rest of one's experience and labeled as having been either recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between recollecting the (contextual) details of an experience and having a general feeling of familiarity~\citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory~\citep[for review see][]{Kaha12}.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep[for review, also see][]{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a naturalistic event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of \textit{exact} recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the essence (or a general summary) of an experience but forget (or neglect to recount) particular low-level details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

How might we formally characterize the \textit{essence} of an experience, and whether it has been recovered by the rememberer?  And how might we distinguish an experience's overarching essence from its low-level details?  One approach is to start by considering some fundamental properties of the dynamics of our experiences.  Each given moment of an experience tends to derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a \textit{context} for each new moment~\citep[e.g.,][]{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep[for review see][]{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we find ourselves in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts~\citep[e.g., when we walk through a doorway;][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect its low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here, we sought to examine how the temporal dynamics of a naturalistic experience were later reflected in participants' memories.  We also sought to leverage the above conceptual insights into the distinctions between an experience's essence and its low-level details to build models that explicitly quantified these distinctions.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television show \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode and of participants' verbal recalls.  Our framework uses topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls by projecting each moment into a word embedding space.  We then use hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast both naturalistic experiences and memories of those experiences as geometric \textit{trajectories} through word embedding space that describe how they evolve over time. Under this framework, successful remembering entails verbally traversing the content trajectory of the episode, thereby reproducing the shape (essence) of the original experience.  Our framework captures the episode's essence in the sequence of geometric coordinates for its events, and its low-level details by examining its within-event geometric properties.

Comparing the overall shapes of the topic trajectories for the episode and participants' recalls reveals which aspects of the episode's essence were preserved (or lost) in the translation into memory.  We also develop two metrics for assessing participants' memories for low-level details: (1) the \textit{precision} with which a participant recounts details about each event, and (2) the \textit{distinctiveness} of their recall for each event, relative to other events.  We examine how these metrics relate to overall memory performance as judged by third-party human annotators.  We also compare and contrast our general approach to studying memory for naturalistic experiences with standard metrics for assessing performance on more traditional memory tasks, such as list-learning.  Last, we leverage our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode and/or how participants would later recount it.

\section*{Results}
To characterize the dynamic content of the \textit{Sherlock} episode and participants' subsequent recountings, we used a topic model~\citep{BleiEtal03} to discover the episode's latent themes.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a \textit{topics matrix} whose rows are \textit{topics} (or latent themes) and whose columns correspond to words in the vocabulary. The entries in the topics matrix reflect how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix} with one row per document and one column per topic.  The topic proportions matrix describes the mixture of discovered topics reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1,000 (manually delineated) time segments spanning the roughly 50 minute video used in their study.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1,976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used detailed, hand-generated annotations describing each manually identified time segment from the episode to fit a topic model.  Three example frames from the episode (first row) are displayed, along with their descriptions from the corresponding episode annotation (second row) and an example participant's recall transcript (third row).  We used the topic model (fit to the episode annotations) to estimate topic vectors for each moment of the episode and each sentence of participants' recalls.  Example topic vectors are displayed in the bottom row (blue: episode annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively), along with the 10 highest-weighted words for each topic.  Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also \textit{sparse}, in that only a small number of topics (typically one or two) tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit \textit{persistence} (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with \textit{occasional rapid changes} (i.e., occasionally topic weights would change abruptly from one timepoint to the next).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by \cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the \textit{event boundaries} where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of events into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods}).  We then created a stable summary of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 episode timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 episode events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the episode.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recall events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of episode events (Panel C) and recall events (from the example participant; Panel F).  For similar plots for all participants, see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of episode events and recall events (across all 17 participants).  To create the figure, each recalled event was assigned to the episode event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions matrix (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  Whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced while viewing the episode, and the events they later recalled.  We labeled each recall event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transitioned between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with early events (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging meaningful narrative associations and references over long timescales throughout the episode.

Clustering scores are often used by memory researchers to characterize how people organize their memories of words on a studied list~\citep[for review, see][]{PolyEtal09}.  We defined analogous measures to characterize how participants organized their memories for episodic events (see \textit{Methods} for details).  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed episode events tended to appear nearby in participants' recall event sequences (mean clustering score: 0.732, SEM: 0.033).  Participants with higher temporal clustering scores tended to exhibit better overall memory for the episode, according to both \cite{ChenEtal17}'s hand-counted numbers of recalled scenes from the episode (Pearson's $r(15) = 0.49,~p = 0.046$) and the numbers of episode events that best-matched at least one recall event (i.e., model-estimated number of events recalled; Pearson's $r(15) = 0.59,~p = 0.013$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar episode events together (mean clustering score: 0.650, SEM: 0.032), and that semantic clustering score was also related to both hand-counted  (Pearson's $r(15) = 0.65,~p = 0.005$) and model-estimated (Pearson's $r(15) = 0.58,~p = 0.015$) numbers of recalled events.

>>>>

\newcommand{\topicopt}{1}
\newcommand{\topics}{2}
\newcommand{\featureimportance}{3}
\newcommand{\corrmats}{5}
\newcommand{\matchmats}{6}
\newcommand{\kopt}{7}
\newcommand{\arrows}{4}

@@

\begin{abstract}
The mental contexts in which we interpret experiences are often person-specific, even when the experiences themselves are shared. We developed a geometric framework for mathematically characterizing the subjective conceptual content of dynamic naturalistic experiences. We model experiences and memories as ``trajectories'' through word embedding spaces whose coordinates reflect the universe of thoughts under consideration. Memory encoding can then be modeled as geometrically preserving or distorting the ``shape'' of the original experience. We applied our approach to data collected as participants watched and verbally recounted a television episode while undergoing functional neuroimaging. Participants’ recountings all preserved coarse spatial properties (essential narrative elements), but not fine spatial scale (low-level) details, of the episode’s trajectory. We also identified networks of brain structures sensitive to these trajectory shapes. Our work provides insights into how we preserve and distort our ongoing experiences when we encode them into episodic memories.
\end{abstract}

\section*{Introduction}
What does it mean to remember something? In traditional episodic memory experiments (e.g., list-learning or trial-based experiments\citep{Murd62a, Kaha96}), remembering is often cast as a discrete, binary operation: each studied item may be separated from the rest of one's experience and labeled as having been either recalled or forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between recollecting the (contextual) details of an experience and having a general feeling of familiarity~\citep{Yone02}. Using well-controlled, trial-based experimental designs, the field has amassed a wealth of information regarding human episodic memory~\citep{Kaha12}.  However, there are fundamental properties of the external world and our memories that trial-based experiments are not well suited to capture~\citep{KoriGold94, HukEtal18}.  First, our experiences and memories are continuous, rather than discrete---isolating a naturalistic event from the context in which it occurs can substantially change its meaning.  Second, whether or not the rememberer has precisely reproduced a specific set of words in describing a given experience is nearly orthogonal to how well they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, the number or proportion of exact recalls is often considered to be a primary metric for assessing the quality of participants' memories.  Third, one might remember the essence (or a general summary) of an experience but forget (or neglect to recount) particular low-level details.  Capturing the essence of what happened is often a main goal of recounting an episodic memory to a listener, whereas the inclusion of specific low-level details is often less pertinent.

How might we formally characterize the ``essence'' of an experience, and whether it has been recovered by the rememberer?  And how might we distinguish an experience's overarching essence from its low-level details?  One approach is to start by considering some fundamental properties of the dynamics of our experiences.  Each given moment of an experience tends to derive meaning from surrounding moments, as well as from longer-range temporal associations~\citep{LernEtal11, Mann19, Mann20}.  Therefore, the timecourse describing how an event unfolds is fundamental to its overall meaning.  Further, this hierarchy formed by our subjective experiences at different timescales defines a context for each new moment~\citep{HowaKaha02a, HowaEtal14}, and plays an important role in how we interpret that moment and remember it later~\citep{MannEtal15, Mann20}.  Our memory systems can leverage these associations to form predictions that help guide our behaviors~\citep{RangRitc12}.  For example, as we navigate the world, the features of our subjective experiences tend to change gradually (e.g., the room or situation we find ourselves in at any given moment is strongly temporally autocorrelated), allowing us to form stable estimates of our current situation and behave accordingly~\citep{ZackEtal07, ZwaaRadv98}.

Occasionally, this gradual drift of our ongoing experience is punctuated by sudden changes, or shifts (e.g., when we walk through a doorway~\citep{RadvZack17}).  Prior research suggests that these sharp transitions (termed ``event boundaries'') help to discretize our experiences (and their mental representations) into ``events''~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distills those experiences down to their essences.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}.  This work also suggests a means of distinguishing the essence of an experience from its low-level details:  The overall structure of events and event transitions reflects how the high-level experience unfolds (i.e., its essence), while subtler event-level properties reflect its low-level details.  Prior research has also implicated a network of brain regions (including the hippocampus and the medial prefrontal cortex) in playing a critical role in transforming experiences into structured and consolidated memories ~\citep{TompDava17}.

Here, we sought to examine how the temporal dynamics of a naturalistic experience were later reflected in participants' memories.  We also sought to leverage the above conceptual insights into the distinctions between an experience's essence and its low-level details to build models that explicitly quantified these distinctions.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television show \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode and of participants' verbal recalls.  Our framework uses topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls by projecting each moment into a word embedding space.  We then use hidden Markov models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast both naturalistic experiences and memories of those experiences as geometric ``trajectories'' through word embedding space that describe how they evolve over time. Under this framework, successful remembering entails verbally traversing the content trajectory of the episode, thereby reproducing the shape (essence) of the original experience.  Our framework captures the episode's essence in the sequence of geometric coordinates for its events, and its low-level details by examining its within-event geometric properties.

Comparing the overall shapes of the topic trajectories for the episode and participants' recalls reveals which aspects of the episode's essence were preserved (or lost) in the translation into memory.  We also develop two metrics for assessing participants' memories for low-level details: (1) the ``precision'' with which a participant recounts details about each event, and (2) the ``distinctiveness'' of their recall for each event, relative to other events.  We examine how these metrics relate to overall memory performance as judged by third-party human annotators.  We also compare and contrast our general approach to studying memory for naturalistic experiences with standard metrics for assessing performance on more traditional memory tasks, such as list-learning.  Last, we leverage our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode and/or how participants would later recount it.

\section*{Results}
To characterize the dynamic content of the \textit{Sherlock} episode and participants' subsequent recountings, we used a topic model~\citep{BleiEtal03} to discover the episode's latent themes.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a ``topics matrix'' whose rows are ``topics'' (or latent themes) and whose columns correspond to words in the vocabulary. The entries in the topics matrix reflect how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a ``topic proportions matrix'' with one row per document and one column per topic.  The topic proportions matrix describes the mixture of discovered topics reflected in each document.

Chen et al. (2017) collected hand-annotated information about each of 1000 (manually delineated) time segments spanning the roughly 50 minute video used in their study~\cite{ChenEtal17}.  Each annotation included: a brief narrative description of what was happening, the location where the action took place, the names of any characters on the screen, and other similar details (for a full list of annotated features, see \textit{Methods}).  We took the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features from all annotations as the vocabulary for the topic model.  We then concatenated the sets of words across all features contained in overlapping sliding windows of (up to) 50 annotations, and treated each window as a single document for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the episode (see \textit{Methods}; Fig.~\ref{fig:schematic}, Supp.\ Fig.~\topics).  We note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of collections of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single ``topic vector'' for each sliding window of annotations transformed by the topic model.  We then stretched (interpolated) the resulting windows-by-topics matrix to match the time series of the 1976 fMRI volumes collected as participants viewed the episode.

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/schematic}
\caption{\small \textbf{Topic weights in episode and recall content.} We used detailed, hand-generated annotations describing each manually identified time segment from the episode to fit a topic model.  Three example frames from the episode (first row) are displayed, along with their descriptions from the corresponding episode annotation (second row) and an example participant's recall transcript (third row).  We used the topic model (fit to the episode annotations) to estimate topic vectors for each moment of the episode and each sentence of participants' recalls.  Example topic vectors are displayed in the bottom row (blue: episode annotations; green: example participant's recalls).  Three topic dimensions are shown (the highest-weighted topics for each of the three example scenes, respectively), along with the 10 highest-weighted words for each topic.  Supplementary Figure~\topics~provides a full list of the top 10 words from each of the discovered topics.}
\label{fig:schematic}
\end{figure}

The 32 topics we found were heavily character-focused (i.e., the top-weighted word in each topic was nearly always a character) and could be roughly divided into themes centered around Sherlock Holmes (the titular character), John Watson (Sherlock's close confidant and assistant), supporting characters (e.g., Inspector Lestrade, Sergeant Donovan, or Sherlock's brother Mycroft), or the interactions between various groupings of these characters (Supp.\ Fig.~\topics).  This likely follows from the frequency with which these terms appeared in the episode annotations.  Several of the identified topics were highly similar, which we hypothesized might allow us to distinguish between subtle narrative differences if the distinctions between those overlapping topics were meaningful.  The topic vectors for each timepoint were also sparse, in that only a small number of topics (typically one or two) tended to be ``active'' in any given timepoint (Fig.~\ref{fig:model}A).  Further, the dynamics of the topic activations appeared to exhibit persistence (i.e., given that a topic was active in one timepoint, it was likely to be active in the following timepoint) along with occasional rapid changes (i.e., occasionally topic weights would change abruptly from one timepoint to the next).  These two properties of the topic dynamics may be seen in the block diagonal structure of the timepoint-by-timepoint correlation matrix (Fig.~\ref{fig:model}B) and reflect the gradual drift and sudden shifts fundamental to the temporal dynamics of many real-world experiences, as well as television episodes.  Given this observation, we adapted an approach devised by Baldassano et al. (2017)~\cite{BaldEtal17}, and used a hidden Markov model (HMM) to identify the ``event boundaries'' where the topic activations changed rapidly (i.e., the boundaries of the blocks in the temporal correlation matrix; event boundaries identified by the HMM are outlined in yellow in Fig.~\ref{fig:model}B).  Part of our model fitting procedure required selecting an appropriate number of events into which the topic trajectory should be segmented.  To accomplish this, we used an optimization procedure that maximized the difference between the topic weights for timepoints within an event versus timepoints across multiple events (see \textit{Methods}).  We then created a stable summary of the content within each episode event by averaging the topic vectors across the timepoints spanned by each event (Fig.~\ref{fig:model}C).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modeling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 episode timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries discovered by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 episode events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the episode.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  For similar plots for all participants, see Supplementary Figure~\corrmats.  \textbf{F.} Average topic vectors for each of the 22 recall events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of episode events (Panel C) and recall events (from the example participant; Panel F).  For similar plots for all participants, see Supplementary Figure~\matchmats.  \textbf{H.} Average correlations between each pair of episode events and recall events (across all 17 participants).  To create the figure, each recalled event was assigned to the episode event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the episode topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the episode (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions matrix (Fig.~\ref{fig:model}E).  As in the episode correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  We used the same HMM-based optimization procedure that we had applied to the episode's topic proportions matrix (see \textit{Methods}) to estimate an analogous set of event boundaries in the participant's recounting of the episode (outlined in yellow).  We carried out this analysis on all 17 participants' recall topic proportions matrices (Supp.\ Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Supp.\ Fig.~\corrmats), each participant appeared to have a unique ``recall resolution,'' reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter-duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the episode with different levels of detail---i.e., some might recount only high-level essential plot details, whereas others might recount low-level details instead (or in addition).  The second clear pattern present in every individual participant's recall correlation matrix was that, unlike in the episode correlation matrix, there were substantial off-diagonal correlations.  One potential explanation for this finding is that the topic models, trained only on episode annotations, do not capture topic proportions in participants' ``held-out'' recalls as accurately.  A second possibility is that, whereas each event in the original episode was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appeared to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event (Fig.~\ref{fig:model}E, Supp.\ Fig.~\corrmats)~\citep{MannEtal11, HowaEtal12, Mann19}.

The above results demonstrate that topic models capture the dynamic conceptual content of the episode and participants' recalls of the episode.  Further, the episode and recalls exhibit event boundaries that can be identified automatically using HMMs to segment the dynamic content.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced while viewing the episode, and the events they later recalled.  We labeled each recall event as matching the episode event with the most similar (i.e., most highly correlated) topic vector (Fig.~\ref{fig:model}G, Supp.\ Fig.~\matchmats).  This yielded a sequence of ``presented'' events from the original episode, and a (potentially differently ordered) sequence of ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first (probability of first recall~\citep{AtkiShif68, PostPhil65, WelcBurn24}; Fig.~\ref{fig:list-learning}A); how participants most often transitioned between recalls of the events as a function of the temporal distance between them (lag-conditional response probability~\citep{Kaha96}; Fig.~\ref{fig:list-learning}B); and which events they were likely to remember overall (serial position recall analyses~\citep{Murd62a}; Fig.~\ref{fig:list-learning}C). Some of the patterns we observed appeared to be similar to classic effects from the list-learning literature.  For example, participants had a higher probability of initiating recall with early events (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). However, unlike what is typically observed in list-learning studies, we did not observe patterns comparable to the primacy or recency serial position effects (Fig.~\ref{fig:list-learning}C).  We hypothesized that participants might be leveraging meaningful narrative associations and references over long timescales throughout the episode.

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the episode. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the episode.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the episode.  All panels: error ribbons denote the bootstrap-estimated 95\% confidence interval.}
  \label{fig:list-learning}
\end{figure}
